%
%
% UCSD Doctoral Dissertation Template
% -----------------------------------
% http:\\ucsd-thesis.googlecode.com
%
%
% ----------------------------------------------------------------------
% WARNING: 
%
%  This template has not endorced by OGS or any other official entity.
%  The official formatting guide can be obtained from OGS.
%  It can be found on the web here:
%  http://ogs.ucsd.edu/AcademicAffairs/Documents/Dissertations_Theses_Formatting_Manual.pdf
%
%  No guaranty is made that this LaTeX class conforms to the official UCSD guidelines.
%  Make sure that you check the final document against the Formatting Manual.
%  
%  That being said, this class has been used successfully for publication of 
%  doctoral theses.  
%
%  The ucsd.cls class files are only valid for doctoral dissertations.
%
%
% ----------------------------------------------------------------------
% GETTING STARTED:
%
% Lots of information can be found on the project wiki:
% http://code.google.com/p/ucsd-thesis/wiki/GettingStarted
%
%
% To make a pdf from this template use the command:
%  pdflatex template
%
%
% To get started please read the comments in this template file 
% and make changes as appropriate.
%
%
% ----------------------------------------------------------------------
%
% A thesis using this template and class file was last successfully 
% submitted on 2009/03/19 (at least as far as I know).
%
% If you successfully submit a thesis with this package please let us
% know.
%
% ----------------------------------------------------------------------
% If you desire more control, please see the attached files:
%
%   * ucsd.cls    -- Class file
%   * uct10.clo   -- Configuration files for font sizes 10pt, 11pt, 12pt
%     uct11.clo                            
%     uct12.clo
%
% ----------------------------------------------------------------------



% Setup the documentclass 
% default options: 11pt, oneside, final
%
% fonts: 10pt, 11pt, 12pt -- are valid for UCSD dissertations.
% sides: oneside, twoside -- note that two-sided theses are not accepted 
%                            by OGS.
% mode: draft, final      -- draft mode switches to single spacing, 
%                            removes hyperlinks, and places a black box
%                            at every overfull hbox (check these before
%                            submission).
% chapterheads            -- Include this if you want your chapters to read:
%                              Chapter 1
%                              Title of Chapter
%
%                            instead of
%
%                              1 Title of Chapter
\documentclass[12pt,chapterheads]{ucsd}



% Include all packages you need here.  
% Some standard options are suggested below.
%
% See the project wiki for information on how to use 
% these packages. Other useful packages are also listed there.
%
%   http://code.google.com/p/ucsd-thesis/wiki/GettingStarted



%% AMS PACKAGES - Chances are you will want some or all 
%    of these if writing a dissertation that includes equations.
%  \usepackage{amsmath, amscd, amssymb, amsthm}

%% GRAPHICX - This is the standard package for 
%    including graphics for latex/pdflatex.
\usepackage{graphicx}

%% from fiverr
\usepackage{booktabs}
\usepackage{tabularx}
\usepackage{array}
\newcolumntype{P}[1]{>{\centering\arraybackslash}p{#1}}
\newcolumntype{M}{>{\centering\arraybackslash} m{2cm} }

\usepackage{hyphenat} % allow for hyphenated words to be further hyphenated
                      % https://tex.stackexchange.com/a/2715
\usepackage{rotating}
\usepackage{multirow}
\usepackage{textcomp}
\usepackage{amssymb}
\usepackage{url}
%% from fiverr

%% SUBFIGURE - Use this to place multiple images in a
%    single figure.  Subfigure will handle placement and
%    proper captioning (e.g. Figure 1.2(a))
% \usepackage{subfigure}

%% LATIN MODERN FONTS (replacements for Computer Modern)
% \usepackage{lmodern}
% \usepackage[T1]{fontenc}

%% INDEX
%   Uncomment the following two lines to create an index: 
\usepackage{makeidx}
\makeindex
%   You will need to uncomment the \printindex line near the
%   bibliography to display the index.  Use the command
% \index{keyword} 
%   within the text to create an entry in the index for keyword.

%% HYPERLINKS
%   To create a PDF with hyperlinks, you need to include the hyperref package.
%   THIS HAS TO BE THE LAST PACKAGE INCLUDED!
%   Note that the options plainpages=false and pdfpagelabels exist
%   to fix indexing associated with having both (ii) and (2) as pages.
%   Also, all links must be black according to OGS.
%   See: http://www.tex.ac.uk/cgi-bin/texfaq2html?label=hyperdupdest
%   Note: This may not work correctly with all DVI viewers (i.e. Yap breaks).
%   NOTE: hyperref will NOT work in draft mode, as noted above.
 \usepackage[colorlinks=true, pdfstartview=FitV, 
             linkcolor=black, citecolor=black, 
             urlcolor=black, plainpages=false,
             pdfpagelabels]{hyperref}
% \hypersetup{ pdfauthor = {Your Name Here}, 
%              pdftitle = {The Title of The Dissertation}, 
%              pdfkeywords = {Keywords for Searching}, 
%              pdfcreator = {pdfLaTeX with hyperref package}, 
%              pdfproducer = {pdfLaTeX} }


\usepackage{verbatim}

\usepackage[toc,nopostdot,style=alttree]{glossaries}
\glssetwidest{PERMANOVAX}% widest name
\makeglossaries
\loadglsentries[main]{abbreviations}

\begin{document}


%% FRONT MATTER
%
%  All of the front matter.
%  This includes the title, degree, dedication, vita, abstract, etc..
\include{frontmatter}


%% DISSERTATION

\chapter{The hidden world within ourselves}\label{chapter_review}

Recent years have seen an increased interest in the study of microbial
communities, partly driven by the affordability of the enabling technologies,
this new focus provides a previously ignored axis of understanding to a vast
number of research fields. In addition, this has uncovered a number of
associations and demonstrated mechanisms, where microbial communities are key
explanatory variables.

In fields where microbial communities were already taken into consideration,
these advancements allowed experiments to be conducted at an unprecedented 
scale \cite{RN35, RN4061, RN4267}. As a consequence the underlying software and 
methods to analyze these studies needed to accommodate volumes of information
previously unseen. Notably, most of the software used in this context has been
developed as a consequence of interdisciplinary experiments that span a broad
range of seemingly disjoint fields (biofuel development \cite{RN4268}, built 
environment \cite{RN4270, RN4083}, forensics \cite{RN4269, RN4271}, etc).  As 
such we motivate the rest of the work in this thesis by presenting a survey of 
human-gut associated finding in the microbiome.  The following section 
appeared in the journal of \textsl{Annual Reviews, 2017}.

\include{chapter_review}

\chapter{Exploratory microbiome data visualization}\label{exploratory_chapter}

As we learnt in Chapter~\ref{chapter_review}, there's a wealth of information 
that can be acquired through the study of microbial communities. The 
acquisition of this knowledge, however, is often only possible through the 
usage of specialized software for analysis and visualization. As noted before, 
\gls{qiime} is a successful\footnote{At the moment of this writing the original 
paper has been cited over 8,000 times according to Google Scholar, and the 
Python package has been downloaded over 34,000 times, data from 
\url{https://pypi.org} and \url{https://anaconda.org}.} microbiome analysis 
pipeline that integrates a collection of bioinformatics tools, and makes these 
programs available through a unified user-interface. Until its fifth release, 
\gls{qiime} relied on \gls{king} to display three-dimensional $\beta$-diversity 
plots. \gls{king} was originally developed as a molecular viewer, and was 
\textit{hacked} to act as a scatter-plot viewer. As datasets got larger and 
more common, it became clear that we needed to develop our own solution, thus 
we created Emperor.

Emperor is an interactive $\beta$-diversity viewer, tailored to fit the 
\gls{qiime} ecosystem. $\beta$-diversity plots, commonly referred to as 
ordinations or dimensionality reductions, are often the starting point when 
analyzing a microbiome dataset. Their ability to overview the relative 
differences of an unlimited number of samples with an unlimited number of 
covariates at a time, makes them useful to diagnose and troubleshoot any 
problems that may have occurred during sample collection, preparation or 
processing. For example, batch effects are often easy to notice in this 
representation\cite{Gibbons165910}, but more challenging to see in 
feature-level analyses, specially if those features specifically are not 
affected by the batch effects.

As use-cases arose, these were integrated as part of the software, something 
that was not previously possible with \gls{king}. A notable feature, was 
allowing users to animate longitudinal sampling schemes, this ability was of 
key importance for the work in Chapter~\ref{chapter_ibds} and 
Chapter~\ref{chapter_fmts}. The following two sections introduce Emperor, and 
further discuss the use of animated ordinations as a way to interact with 
microbiome data. Section~\ref{section_emperor} was published in the journal 
\textsl{GigaScience, 2013} and Section~\ref{section_animations} was published 
in the journal \textsl{Cell Host \& Microbe, 2017}.

\include{chapter_EMPeror}
\include{chapter_animations}

\chapter{Inflammatory bowel disease in Dogs and Humans}

\gls{ibd} is a family of conditions divided into two major categories, \gls{uc} 
and \gls{cd}, each of which can be divided into other subtypes specifying 
location, age of diagnosis, severity and behaviour of the disease 
\cite{RN4265}.  In all cases \gls{ibd} is associated with diarrhea, 
inflammation and flaring episodes of exacerbated discomfort. In general, 
diagnostic methods rely on questionnaires, direct examinations of affected 
intestinal tissue, or quantification of calprotectin \cite{Sipponen2008}.

Our understanding of the community dynamics that take place in subjects with 
\gls{ibd} is extensive, though regrettably not sufficient. One major aspect 
that has yet to be explored is the inter-species similarities of their 
microbial communities. Making sense out of the interactions that take place in 
otherwise different systems, could potentially lead the way to fully comprehend 
the exogenous and endogenous factors that cause and regulate this family of 
conditions.

In this chapter, we introduce work motivated by our recent efforts at 
characterizing the microbiome in new-onset and treatment-naive Crohn's disease 
subjects \cite{RN154}. Specifically, we focus on comparing the microbiomes of 
\gls{ibd} affected humans and dogs. As a result, we further expanded the 
repertoire of features that we know and associate with \gls{ibd} in dogs 
\cite{RN153}, but most importantly we use this new information and through a 
meta-analysis, we contextualize it with data collected from humans 
\cite{RN154}.  We find that the dysbiosis networks\footnote{A dysbiosis network 
is the graph resulting from using a correlation matrix as a weighted adjacency 
matrix.} are overlapping, both in structure and in the microbes that are 
associated with the \gls{ibd} and non-\gls{ibd} affected communities, however 
we uncover a taxon that shifts from being non \gls{ibd}-associated in humans to 
being \gls{ibd}-associated in dogs.  Section~\ref{dogs} appeared in the journal 
\textsl{Nature Microbiology, 2016}.

\include{chapter_dogs}

\chapter{Dynamic features of inflammatory bowel disease}\label{chapter_ibds}

In recent years, clinical research in \gls{ibd} has benefited from multi-'omic 
(genomic\cite{RN4217}, metabolomic\cite{RN4264} and metagenomic\cite{RN4263}) 
characterizations of the condition. Each describing a new component of 
relationships, that one by one, seem to uncover the processes regulating the 
inflammation and well-being of the affected hosts. Nonetheless longitudinal 
descriptions of the disease seemed to be lagging behind.

This chapter introduces two of the pioneering longitudinal studies of 
\gls{ibd}, one where the sampling is sparse (every three months) and another 
where the sampling is dense (as much as every day). Both cohorts present 
increased rates of microbial variation, specifically we see that this appears 
to be different for the different subtypes of \gls{ibd}. In order to visualize 
the increased variation, we rely on the techniques introduced in 
Chapter~\ref{exploratory_chapter}, and on the creation of a reference plane 
that acts as a proxy for healthy variation.

Motivated by these findings, we use a classification model to determine the 
benefits of increased longitudinal sampling. For both (independently collected 
and sequenced) datasets, we observe that including more than one sample per 
subject increases the performance of a classifier. This property is only 
possible when we create per-subject consensus features based on multiple 
samples.

Although we exercise and test this method with human-gut data, the approach is 
not restricted in any way to this environment, other applications can benefit, 
specifically where dynamic descriptors might be predictive of a state of 
interest.

% TODO: add publication
Section~\ref{plane} appeared in \textsl{Nature Microbiology, 2017}, and 
Section~\ref{ibd} appeared in \dots

\include{chapter_plane}
\include{chapter_ibd}

\chapter{Restoring a lost ecosystem}\label{chapter_fmts}
\glsresetall

\gls{cdi} is a hospital-acquired infection resulting from antibiotic 
administration for an unrelated condition. Paradoxically, the treatment for 
\gls{cdi} is to prescribe more antibiotics. This treatment translates in
depleted microbial diversity, as a consequence, resources that could be 
consumed by other communities are now available for \textit{Clostridium 
difficile} to thrive on.  More recently \gls{cdi} has been treated through the 
administration of \glspl{fmt}, with a success rate above 90\%\cite{RN4129}. A 
\gls{fmt} consists in reintroducing bacteria into the \gls{gi} tract of an 
affected patient. Frequently, after a day, the symptoms of \gls{cdi} disappear 
and the recipients recover. In this chapter, we examine two cohorts and focus 
on the short and long-term changes in the gut microbiota after a \gls{fmt}.

In the first cohort and relying on an animated $\beta$-diversity plot (as 
introduced in Section~\ref{section_animations}), we visualize the immediate 
changes in the gut microbiome after a \gls{fmt}. In addition, we also visualize 
the stability of the microbiome after the transplant, and note that the only 
point of instability occurs right after the \gls{fmt}.

The second cohort is composed by patients suffering from \gls{cdi} and in some 
cases with underlying \gls{ibd}. We find that, while both groups recover from 
the \gls{cdi}, the micro-ecological effects of \gls{fmt} appear to be dampened 
in subjects with underlying \gls{ibd}. Overall, individual phylogenetic 
diversity is lower, and the number of taxa differentially present before and 
after the transplant is also smaller (when compared to the non-\gls{ibd} 
counterparts).

Both sections (Section~\ref{section_moviefmt} and Section~\ref{section_fmt}) 
were published in the journal \textsl{Microbiome} in \textsl{2015} and 
\textsl{2017} respectively.

\include{chapter_moviefmt}
\include{chapter_fmt}

\chapter{Conclusions}

\include{chapter_conclusions}

%\appendix
%\chapter{Final notes}
%  Remove me in case of abdominal pain.



%% END MATTER
\printindex %% Uncomment to display the index
% \nocite{}  %% Put any references that you want to include in the bib 
%               but haven't cited in the braces.
% \bibliographystyle{alpha}  %% This is just my personal favorite style. 
%                              There are many others.
\bibliography{references}
\bibliographystyle{plain}

\end{document}
